Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-22 analysis/paper/article.qmd
minor edit
More text added on valorization of tenure-track jobs in higher ed.

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, the wiki is a set of freely accessible web pages that anyone can edit anonymously or with a free user account. The site is hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from a variety of sources such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx [ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

Demand for jobs focusing on the Americas has been generally high over the past decade, with a peak in 2019-2020 and a subsequent decrease. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking in 2016 and experiencing a nadir in 2019, showing an inverse pattern to the Americas. Asia and India, the Near East, and Europe <!--crying face emoji BM: sorry! --> rarely occur as a geographical focus in job ads at US institutions. Asia and India, Africa, and the Americas appear correlated with each other, while the Near East and Mediterranean are inversely correlated in an opposite trend.

@@

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing "the last good job[s] in America" [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–20000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Job Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore "a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline" [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the "maladaptive" consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This suggests that these patterns are a reflection of the wiki, rather than the market itself. <!--JB MAYBE ADD A NOTE HERE FROM A CRISIS PAPER, PERHAPS MELISSA'S, ABOUT DISPARITY IN TRAINING AND APPLIED ANTHROPOLOGY. BM: [inserting earplugs] GO FOR IT! -->

The downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

@@

Our data suggest that archaeology faculty at US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. Due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity, however,it is illegal in the US for universities use race or ethnicity as a primary factor in hiring. As a result, universities appear to have tailored the content of their job ads to focus on topics where they expected Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019] <!--Should we cite Cate Frieman here? BM: grimace emoji -->. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

>>>>

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, the wiki is a set of freely accessible web pages that anyone can edit anonymously or with a free user account. The site is hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from a variety of sources such as *The* *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx \[ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug\] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

Demand for jobs focusing on the Americas has been generally high over the past decade, with a peak in 2019-2020 and a subsequent decrease. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking in 2016 and experiencing a nadir in 2019, showing an inverse pattern to the Americas. Asia and India, the Near East, and Europe rarely occur as a geographical focus in job ads at US institutions. Asia and India, Africa, and the Americas appear correlated with each other, while the Near East and Mediterranean are inversely correlated in an opposite trend.

@@

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. The wiki’s emphasis on tenure-track jobs runs counter to well-established patterns in the academic job market, wherein short-term and contingent positions are fast outpacing "the last good job\[s\] in America" [@aronowitz2001]. Research across anthropological sub-fields has shown that the number of tenure-track positions advertised within the discipline has also declined. Analyzing the job ads posted in *Anthropology News* in 1999–20000 versus 2019–2020, for example, Gershon and Rachok [-@HelloTristesTropes2021] found that the number of tenure-track positions decreased by 25% between their two time slices. Similarly, in his examination of jobs ads posted on the Biological Anthropology Academic Job Wiki between 2010–2017, Passalacqua noted that approximately one quarter of the 474 ads posted were for adjunct or visiting staff [@Passalacqua_2018, pp.773]. Though Passalacqua did not examine how the proportion of job types changed over time, he did underscore "a current diverging trend of decreasing academic job advertisements and increasing doctoral degrees in biological anthropology which could result in serious consequences for the discipline" [-@Passalacqua_2018, pp.773]. Such observations have a surprising longevity. Rogge raised the alarm nearly half a century ago, warning of the "maladaptive" consequences of the exponential rate of disciplinary growth and impending consequences for job seekers [@rogge1976, pp.839]. The dwindling number of permanent academic positions is not a trend unique to anthropology—as of 2019, the AAUP documented a 36% increase in contingent positions over the preceding 15 years (AAUP 2022), and as of 2022, contingent positions make up more than half of faculty positions in the United States [@colby2022annual] .

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This congruence between subfields suggests that these patterns are a reflection of the *wiki*, rather than the market itself.

It is no secret that tenure-track jobs have traditionally been considered the "gold standard" of career outcomes for PhD students in the social sciences and humanities. As Kelsky outlines in *The Professor Is In*, her best-selling guide for PhD students pursuing faculty careers, "...most ranking graduate programs sill consider any Ph.D. who doesn't land a tenure-track job a failure or an aberration... Graduate students absorb this value system and judge themselves harshly" [-@kelsky2015, pp. 11]. These rubrics of success have been common across the humanities and social sciences for over half a century. Evaluating the demise of the National Endowment for Humanities Funded "Program to ready PhDs for Careers in Business" (CIB), which aimed to facilitate humanities PhDs' transition to the corporate world in the late 1970s, Franczak points to the dissonance between participants' incentives and outcomes:

“The biggest lesson we can take from CIB is also the most obvious. PhDs, especially in the humanities, want to be academics. The deep reservoir of adjunct or contingent faculty that elite and non-elite universities alike depend on for their courses is testament to this fact—as is the excellent scholarship so many adjuncts produce without department support. To pretend otherwise is disingenuous, and, as CIB shows, possibly dangerous, too" [@bessner2021, pp. 36].

The desire to pursue an academic career also characterizes more recent PhD cohorts, including those within archaeology. Brami et al's survey found that despite the long odds and grueling precaraity, 71% of their early career respondents wished to remain in academia [=@bramiPrecariousFutureReflections2023, pp. 242]. Overall, the obsessive valorization of permanent faculty positions in the face of a precipitous decline in available jobs is indicative of the potent combination of denial and delusion inacross academia. As Cefkin and Schwegler argue "This asymmetry between the “ideal” career path and the experiences of most graduates highlights the deep and growing chasm between the future that higher education institutions envision—and strongly incentivize—and what is actually happening" [@cefkin].

Returning to our results, the downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

@@

Our data suggest that archaeology faculty at US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. Due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity, however, it is illegal in the US for universities use race or ethnicity as a primary factor in hiring. As a result, universities appear to have tailored the content of their job ads to focus on topics where they expected Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019]. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.
